---
title: "STAT 628 Credit Risk Project: Installment 2 (technical)"
author: "Bohyoon Lee and Salsabila Mahdi"
date: last-modified
format: pdf
geometry:
  - margin=1in
---

```{r}
#| label: setup
#| echo: false
#| warning: false

rm(list = ls())

library(dplyr)
library(ModelMetrics)
library(car)
library(glmnet)
library(ggplot2)
library(knitr)
library(gtsummary)
```

## Data preprocessing and cleaning

First, we make NAICS, WomanOwned, and FICO categorical variables.

```{r load_clean}
installment2_id01 <- read.csv("installment2_id01.csv")

installment2_id01 <- installment2_id01 %>%
  mutate(NAICS = as.factor(substr(NAICS, 1, 2)),
         WomanOwned = as.factor(WomanOwned),
         CorpStructure = as.factor(CorpStructure),
         FICO = as.factor(case_when(
           (300 <= FICO)&(FICO <= 579) ~ "Poor",
           (580 <= FICO)&(FICO <= 669) ~ "Fair",
           (670 <= FICO)&(FICO <= 739) ~ "Good",
           (740 <= FICO)&(FICO <= 799) ~ "Very Good",
           (800 <= FICO)&(FICO <= 850) ~ "Excellent",
         )))

get_mode <- function(categorical_x) {
  mode <- names(which.max(table(categorical_x)))[1]
}

installment2_id01$FICO <- relevel(installment2_id01$FICO, ref = "Good")
installment2_id01$CorpStructure <- relevel(installment2_id01$CorpStructure,
                                           ref = get_mode(installment2_id01$CorpStructure))
installment2_id01$WomanOwned <- relevel(installment2_id01$WomanOwned,
                                        ref = get_mode(installment2_id01$WomanOwned))

installment2_id01 %>% tbl_summary(
    statistic = list(
      all_continuous() ~ c(
        "{median} ({p25}, {p75})",
        "{min}, {max}"
      )
    ),
    type = all_continuous() ~ "continuous2"
  ) %>% 
  modify_header(all_stat_cols() ~ "**N = {N}**") %>% 
  modify_footnote(everything() ~ NA_character_)
# TODO: meaningful units?
```

We then notice a few points have PRSM values that are out of distribution (less than 0 or greater than 2) and remove them.

```{r remove_outliers, warning=FALSE}
# remove outliers
installment2_id01_clean <- subset(installment2_id01, PRSM > 0 & PRSM < 2)
installment2_id01_summary <- installment2_id01_clean %>% 
  summarise(`min PRSM` = min(PRSM),
            `median PRSM` = median(PRSM),
            `mean PRSM` = mean(PRSM),
            `max PRSM` = max(PRSM))

kable(installment2_id01_summary, digits = 3)
```

Above is the summary of the PRSM after removing the outliers.

## Model Development and Selection

### Linearity and predictor transformation

```{r transform, warning=FALSE}
crPlots(lm(PRSM ~ Months, data = installment2_id01_clean))

# Transformations from boxTidwell
boxTidwell_results <- boxTidwell(PRSM ~ TotalAmtOwed +
                                   Volume + Stress + Num_Delinquent +
                                   Num_CreditLines + Months,
          data = installment2_id01_clean)

boxTidwell_results$result[,c("MLE of lambda","Pr(>|t|)" )] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)

lambda_months <- boxTidwell_results$result[c("Months"),c("MLE of lambda")]
lambda_num_delinquent <- boxTidwell_results$result[c("Num_Delinquent"),c("MLE of lambda")]

transformed_model <- lm(PRSM ~ . - Months + I(Months^(lambda_months)) -
                          Num_Delinquent + I(Num_Delinquent^(lambda_num_delinquent)),
                       data = installment2_id01_clean)

summary(transformed_model)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)

crPlots(lm(PRSM ~ I(Months^(lambda_months)), data = installment2_id01_clean))
```

We assess the linearity of the relationship between PRSM and the predictors using component-residual plots and find that Months shows non-linearity in a somewhat logarithmic form. We transform Months to approximately its inverse (to be precise, Box-Tidwell suggests `r round(lambda_months, 2)`). The transformed Month predictor is statistically significant but Num_Delinquent is not.

### Variable and Interaction Selection

We aim to create a parsimonious model that explains the most variance. We do this in a multi-stage selection process.

```{r rm_var, warning=FALSE}
# Step Model
step_model <- step(transformed_model, trace = 0, direction = "both")
summary(step_model)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)
```

First, we use bidirectional stepwise on the model containing all main effects (with Months transformed). It identifies Num_Delinquent, and Num_CreditLines as statistically non-significant predictors and removes them. We also remove Volume as it is statistically non-significant and its effect is most likely covered by Stress.

Secondly, we search for statistically significant two-way interactions with LASSO, which affect PRSM meaningfully. We add quite a few interactions between FICO, Stress, WomanOwned and CorpStructure with other predictors to the model.

```{r interaction_lasso, warning=FALSE}
# Find interactions using LASSO
installment2_id01_clean$Months_transformed =
  (installment2_id01_clean$Months)^(lambda_months)
x <- model.matrix(PRSM ~ .^2, 
                  data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed", "NAICS")])[, -1]
y <- installment2_id01_clean$PRSM
cv_lasso_interactions <- cv.glmnet(x, y, alpha = 1, nfolds = 10)

best_lambda <- cv_lasso_interactions$lambda.min
lasso_coeffs <- coef(cv_lasso_interactions, s = best_lambda)

threshold <- 0.001
significant_terms <- lasso_coeffs[abs(lasso_coeffs[,1]) > threshold, , drop = FALSE]
print(significant_terms)

# potential interactions: FICO:Stress, FICO:WomanOwned, FICO:CorpStructure, 
# FICO:Months_transformed, Stress:WomanOwned, Stress:CorpStructure, 
# WomanOwned:CorpStructure, WomanOwned:Months_transformed, 
# CorpStructure:Months_transformed, FICO:NAICS, WomanOwned:NAICS, 
# CorpStructure:NAICS, Months_transformed:NAICS, Stress:NAICS

interact_model = lm(PRSM ~. + FICO:Stress + FICO:WomanOwned + FICO:CorpStructure 
                    + FICO:Months_transformed + Stress:WomanOwned 
                    + Stress:CorpStructure + WomanOwned:CorpStructure 
                    + WomanOwned:Months_transformed + CorpStructure:Months_transformed
                    + FICO:NAICS + WomanOwned:NAICS + CorpStructure:NAICS 
                    + Months_transformed:NAICS + Stress:NAICS
                      , data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed", "NAICS")])

# printed R summary of the fitted model
step_interact_model = step(interact_model, trace = 0, direction = "both")
summary(step_interact_model)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)

# NAICS is not statistically significant; remove NAICS
interact_model_final = lm(PRSM ~. + FICO:Stress + FICO:WomanOwned + FICO:CorpStructure 
                    + FICO:Months_transformed + Stress:WomanOwned 
                    + Stress:CorpStructure + WomanOwned:CorpStructure 
                    + WomanOwned:Months_transformed + CorpStructure:Months_transformed
                      , data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed")])
step_interact_final = step(interact_model_final, trace = 0, direction = "both")
```

Finally, we test the validity of the model, with variables properly casted and outliers removed from the data, Months transformed, selected main effects, and interaction terms. Normality and homogenity assumptions hold, and most of the predictors have no or moderate correlations with each other. The model also has the lowest RMSE so we choose it as the final model.

```{r final_model_checks, warning=FALSE}
# CHECK ASSUMPTIONS FOR FINAL MODEL
plot(step_interact_final, 2)
plot(step_interact_final, 1)
vif(step_interact_final) # <5, still ok for multicol

# TODO: why ggplot? => A: artefact of exploration phase, we kept it so we'd remember how to make similar plots
ggplot(data = installment2_id01_clean, aes(x = WomanOwned)) +
  geom_bar() +
  facet_wrap(~ FICO)
```

Below is the summary of the fitted model and the RMSE. 
 
```{r summary_fitted}
# Printed R summary of the fitted model. This is the final model
summary(step_interact_final)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.3f", `Pr(>|t|)`))) %>%
  kable(digits = 3)

rmse(step_interact_final$fitted.values, installment2_id01_clean$PRSM) # RMSE: 0.09830548
```



### Interpretation

Main effects in the final model:

1.    **FICO Score**. Compared to borrowers with a **Good** FICO score
    (the baseline), borrowers in other categories perform differently.
    Borrowers with an **Excellent** score are expected to have a higher
    PRSM by `r round(summary(step_interact_final)$coefficients[c("FICOExcellent"),c("Estimate")], 3)`,
    while those with a **Very Good** score see an increase of
    `r round(summary(step_interact_final)$coefficients[c("FICOVery Good"),c("Estimate")], 3)`.
    Conversely, those with a **Fair** score have a lower PRSM by
    `r round(summary(step_interact_final)$coefficients[c("FICOFair"),c("Estimate")], 3)`.
    The negative effect for the "Poor" category is not statistically significant.

2.  **Total Amount Owed**. Total Amount Owed has no practical effect on
    PRSM although it is statistically significant with a coefficient of
    `r round(summary(step_interact_final)$coefficients[c("TotalAmtOwed"),c("Estimate")], 7)`.

3.  **Stress**. Stress has a significantly strong positive effect on PRSM
    with a coefficient of
    `r round(summary(step_interact_final)$coefficients[c("Stress"),c("Estimate")], 3)`.
    For every 0.1 unit increase of Stress, the PRSM increases by
    `r round(summary(step_interact_final)$coefficients[c("Stress"),c("Estimate")], 3)/10`.

4.  **Woman Owned**. Compared to businesses not owned by women, women-owned
    businesses are expected to increase PRSM by
    `r round(summary(step_interact_final)$coefficients[c("WomanOwned1"),c("Estimate")], 3)`.

5.  **Corporate Structure**. Compared to businesses structured as an **LLC**
    (the baseline), all other structures are associated with a lower PRSM.
    **Corporations** are associated with a PRSM that is lower by
    `r round(summary(step_interact_final)$coefficients[c("CorpStructureCorp"),c("Estimate")], 3)`,
    **Partnerships** by
    `r round(summary(step_interact_final)$coefficients[c("CorpStructurePartner"),c("Estimate")], 3)`,
    and **Sole Proprietorships** by
    `r round(summary(step_interact_final)$coefficients[c("CorpStructureSole"),c("Estimate")], 3)`.

6.  **Months in Business**. The transformed Months variable
    (Months^(`r lambda_months`)) gets smaller as the months a business
    is open increases. Its negative coefficient of
    `r round(summary(step_interact_final)$coefficients[c("Months_transformed"),c("Estimate")], 3)`
    indicates that PRSM increases as the business ages, though this effect
    diminishes over time.

Interaction effects in the final model:

1.  **FICO:WomanOwned**. The positive main effect of a business being
    woman-owned is modified by the borrower's FICO score. For a
    woman-owned business with an **Excellent** FICO score, the PRSM
    increases by an *additional*
    `r round(summary(step_interact_final)$coefficients[c("FICOExcellent:WomanOwned1"),c("Estimate")], 3)`.
    Conversely, the effect is reduced for those with **Fair** and **Poor**
    scores, with an additional change of
    `r round(summary(step_interact_final)$coefficients[c("FICOFair:WomanOwned1"),c("Estimate")], 3)`
    and
    `r round(summary(step_interact_final)$coefficients[c("FICOPoor:WomanOwned1"),c("Estimate")], 3)`,
    respectively. The interaction for the "Very Good" category is not
    statistically significant.

2.  **FICO:Months_transformed**. The effect of business age on PRSM is
    significantly different for borrowers in the **Excellent** and
    **Very Good** FICO categories compared to the "Good" baseline. The
    effect of `Months_transformed` changes by an additional
    `r round(summary(step_interact_final)$coefficients[c("FICOExcellent:Months_transformed"),c("Estimate")], 3)`
    for "Excellent" borrowers and by
    `r round(summary(step_interact_final)$coefficients[c("FICOVery Good:Months_transformed"),c("Estimate")], 3)`
    for "Very Good" borrowers. Other FICO interaction terms for this
    variable are not statistically significant.

3.  **Stress:WomanOwned**. The positive effect of Stress on PRSM is even
    stronger for woman-owned businesses. For every 0.1 unit increase in
    Stress, the PRSM for a woman-owned business increases by an
    *additional*
    `r round(summary(step_interact_final)$coefficients[c("Stress:WomanOwned1"),c("Estimate")]*0.1, 3)`
    compared to a non-woman-owned business.

4.  **WomanOwned:CorpStructure**. For woman-owned businesses, the negative
    main effects associated with non-LLC structures are lessened. The
    PRSM for a **woman-owned Corporation** increases by an additional
    `r round(summary(step_interact_final)$coefficients[c("WomanOwned1:CorpStructureCorp"),c("Estimate")], 3)`,
    a **woman-owned Partnership** by
    `r round(summary(step_interact_final)$coefficients[c("WomanOwned1:CorpStructurePartner"),c("Estimate")], 3)`,
    and a **woman-owned Sole Proprietorship** by
    `r round(summary(step_interact_final)$coefficients[c("WomanOwned1:CorpStructureSole"),c("Estimate")], 3)`.

5.  **CorpStructure:Months_transformed**. The effect of business age on
    PRSM is significantly different for corporations compared to LLCs.
    For a **Corporation**, the effect of `Months_transformed` changes by an
    additional
    `r round(summary(step_interact_final)$coefficients[c("CorpStructureCorp:Months_transformed"),c("Estimate")], 3)`.
    The effects for other structures are not statistically significant.



### Baseline Borrower

```{r baseline}
newdata<-installment2_id01_clean[1,]; print(newdata)
as.data.frame(round(predict(step_interact_final, 
        newdata=newdata), 8))
newdata[,c("FICO")] <- "Good"
newdata[,c("TotalAmtOwed")] <- median(installment2_id01_clean$TotalAmtOwed)
newdata[,c("Stress")] <- median(installment2_id01_clean$Stress)
newdata[,c("WomanOwned")] <- "0"
newdata[,c("CorpStructure")] <- "LLC"
newdata[,c("Months")] <- median(installment2_id01_clean$Months)
newdata[,c("Months_transformed")] <- median(installment2_id01_clean$Months)^lambda_months
print(newdata)
as.data.frame(round(predict(step_interact_final, 
        newdata=newdata), 8))
```

The baseline borrower has a Good FICO score, `r median(installment2_id01_clean$TotalAmtOwed)` of Total Amount Owed, and `r median(installment2_id01_clean$Stress)` of Stress, is not a woman, is structured as a LLC, and has opened the business for `r median(installment2_id01_clean$Months)` months.
