---
title: "STAT 628 Credit Risk Project: Installment 2 (technical)"
author: "Bohyoon Lee and Salsabila Mahdi"
date: last-modified
format: pdf
geometry:
  - margin=1in
---

```{r}
#| label: setup
#| echo: false
#| warning: false

rm(list = ls())

library(dplyr)
library(ModelMetrics)
library(car)
library(glmnet)
library(ggplot2)
library(knitr)
library(gtsummary)
```

## Data preprocessing and cleaning

First, we make NAICS, WomanOwned, and FICO categorical variables.

```{r load_clean}
installment2_id01 <- read.csv("installment2_id01.csv")

installment2_id01 <- installment2_id01 %>%
  mutate(NAICS = as.factor(substr(NAICS, 1, 2)),
         WomanOwned = as.factor(WomanOwned),
         FICO = case_when(
           (300 <= FICO)&(FICO <= 579) ~ "Poor",
           (580 <= FICO)&(FICO <= 669) ~ "Fair",
           (670 <= FICO)&(FICO <= 739) ~ "Good",
           (740 <= FICO)&(FICO <= 799) ~ "Very Good",
           (800 <= FICO)&(FICO <= 850) ~ "Excellent",
         ))


installment2_id01 %>% tbl_summary(
    statistic = list(
      all_continuous() ~ c(
        "{median} ({p25}, {p75})",
        "{min}, {max}"
      )
    ),
    type = all_continuous() ~ "continuous2"
  ) %>% 
  modify_header(all_stat_cols() ~ "**N = {N}**") %>% 
  modify_footnote(everything() ~ NA_character_)
# TODO: meaningful units?
```

We then notice a few points have PRSM values that are out of distribution (less than 0 or greater than 2) and remove them.

```{r remove_outliers, warning=FALSE}
# remove outliers
installment2_id01_clean <- subset(installment2_id01, PRSM > 0 & PRSM < 2)
installment2_id01_summary <- installment2_id01_clean %>% 
  summarise(`min PRSM` = min(PRSM),
            `median PRSM` = median(PRSM),
            `mean PRSM` = mean(PRSM),
            `max PRSM` = max(PRSM))

kable(installment2_id01_summary, digits = 3)
```

Above is the summary of the PRSM after removing the outliers.

## Model Development and Selection

### Linearity and predictor transformation

```{r transform, warning=FALSE}
crPlots(lm(PRSM ~ Months, data = installment2_id01_clean))

# Transformations from boxTidwell
boxTidwell_results <- boxTidwell(PRSM ~ TotalAmtOwed +
                                   Volume + Stress + Num_Delinquent +
                                   Num_CreditLines + Months,
          data = installment2_id01_clean)

boxTidwell_results$result[,c("MLE of lambda","Pr(>|t|)" )] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)

lambda_months <- boxTidwell_results$result[c("Months"),c("MLE of lambda")]
lambda_num_delinquent <- boxTidwell_results$result[c("Num_Delinquent"),c("MLE of lambda")]

transformed_model <- lm(PRSM ~ . - Months + I(Months^(lambda_months)) -
                          Num_Delinquent + I(Num_Delinquent^(lambda_num_delinquent)),
                       data = installment2_id01_clean)

summary(transformed_model)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)

crPlots(lm(PRSM ~ I(Months^(lambda_months)), data = installment2_id01_clean))
```

We assess the linearity of the relationship between PRSM and the predictors using component-residual plots and find that Months shows non-linearity in a somewhat logarithmic form. We transform Months to approximately its inverse (to be precise, Box-Tidwell suggests `r round(lambda_months, 2)`). The transformed Month predictor is statistically significant but Num_Delinquent is not.

### Variable and Interaction Selection

We aim to create a parsimonious model that explains the most variance. We do this in a multi-stage selection process.

```{r rm_var, warning=FALSE}
# Step Model
step_model <- step(transformed_model, trace = 0, direction = "both")
summary(step_model)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)
```

First, we use bidirectional stepwise on the model containing all main effects (with Months transformed). It identifies Num_Delinquent, and Num_CreditLines as statistically non-significant predictors and removes them. We also remove Volume as it is statistically non-significant and its effect is most likely covered by Stress.

Secondly, we search for statistically significant two-way interactions with LASSO, which affect PRSM meaningfully. We add quite a few interactions between FICO, Stress, WomanOwned and CorpStructure with other predictors to the model.

```{r interaction_lasso, warning=FALSE}
# Find interactions using LASSO
installment2_id01_clean$Months_transformed =
  (installment2_id01_clean$Months)^(lambda_months)
x <- model.matrix(PRSM ~ .^2, 
                  data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed", "NAICS")])[, -1]
y <- installment2_id01_clean$PRSM
cv_lasso_interactions <- cv.glmnet(x, y, alpha = 1, nfolds = 10)

best_lambda <- cv_lasso_interactions$lambda.min
lasso_coeffs <- coef(cv_lasso_interactions, s = best_lambda)

threshold <- 0.001
significant_terms <- lasso_coeffs[abs(lasso_coeffs[,1]) > threshold, , drop = FALSE]
print(significant_terms)

# potential interactions: FICO:Stress, FICO:WomanOwned, FICO:CorpStructure, 
# FICO:Months_transformed, Stress:WomanOwned, Stress:CorpStructure, 
# WomanOwned:CorpStructure, WomanOwned:Months_transformed, 
# CorpStructure:Months_transformed, FICO:NAICS, WomanOwned:NAICS, 
# CorpStructure:NAICS, Months_transformed:NAICS, Stress:NAICS

interact_model = lm(PRSM ~. + FICO:Stress + FICO:WomanOwned + FICO:CorpStructure 
                    + FICO:Months_transformed + Stress:WomanOwned 
                    + Stress:CorpStructure + WomanOwned:CorpStructure 
                    + WomanOwned:Months_transformed + CorpStructure:Months_transformed
                    + FICO:NAICS + WomanOwned:NAICS + CorpStructure:NAICS 
                    + Months_transformed:NAICS + Stress:NAICS
                      , data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed", "NAICS")])

# printed R summary of the fitted model
step_interact_model = step(interact_model, trace = 0, direction = "both")
summary(step_interact_model)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.2f", `Pr(>|t|)`))) %>%
  kable(digits = 2)

# NAICS is not statistically significant; remove NAICS
interact_model_final = lm(PRSM ~. + FICO:Stress + FICO:WomanOwned + FICO:CorpStructure 
                    + FICO:Months_transformed + Stress:WomanOwned 
                    + Stress:CorpStructure + WomanOwned:CorpStructure 
                    + WomanOwned:Months_transformed + CorpStructure:Months_transformed
                      , data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed")])
step_interact_final = step(interact_model_final, trace = 0, direction = "both")
```

Finally, we test the validity of the model, with variables properly casted and outliers removed from the data, Months transformed, selected main effects, and interaction terms. We find the model to meet the assumptions and have the lowest RMSE so we choose it as the final model. Below is the summary of the fitted model and the RMSE.

```{r summary_fitted}
# Printed R summary of the fitted model. This is the final model
summary(step_interact_final)$coefficients[,c("Estimate","Pr(>|t|)")] %>% 
  as.data.frame %>% 
  mutate(`Pr(>|t|)` = ifelse(`Pr(>|t|)` < 0.05, "<0.05", 
                             sprintf("%.3f", `Pr(>|t|)`))) %>%
  kable(digits = 3)

rmse(step_interact_final$fitted.values, installment2_id01_clean$PRSM) # RMSE: 0.09830548
```

```{r final_model_checks, warning=FALSE}
# CHECK ASSUMPTIONS FOR FINAL MODEL
plot(step_interact_final, 2)
plot(step_interact_final, 1)
vif(step_interact_final) # <5, still ok for multicol

# TODO: why ggplot?
ggplot(data = installment2_id01_clean, aes(x = WomanOwned)) +
  geom_bar() +
  facet_wrap(~ FICO)
```

Normality assumption holds, and most of the predictors have no or moderate correlations with each other.

Interpretation of main effects in the final model:

1.  Compared to borrowers from the category of FICO Excellent, all other borrowers (from the remaining FICO categories) perform worse. FICO Poor effects PRSM by `r round(summary(step_interact_final)$coefficients[c("FICOPoor"),c("Estimate")], 3)`. FICO Fair effects PRSM by `r round(summary(step_interact_final)$coefficients[c("FICOFair"),c("Estimate")], 3)`. FICO Good effects PRSM by `r round(summary(step_interact_final)$coefficients[c("FICOGood"),c("Estimate")], 3)`. FICO Very Good effects PRSM by `r round(summary(step_interact_final)$coefficients[c("FICOVery Good"),c("Estimate")], 3)`.
2.  Total Amount Owed has no practical effect on PRSM although it is statistically significant with the coefficient of `r round(summary(step_interact_final)$coefficients[c("TotalAmtOwed"),c("Estimate")], 7)`.
3.    Stress has a significantly strong positive effect on PRSM with the coefficient of `r round(summary(step_interact_final)$coefficients[c("Stress"),c("Estimate")], 3)`. For every 0.1 unit increase of Stress, the PRSM increases by `r round(summary(step_interact_final)$coefficients[c("Stress"),c("Estimate")], 3)/10`
4.    Compared to businesses not owned by women, women owned businesses are expected to increase PRSM by `r round(summary(step_interact_final)$coefficients[c("WomanOwned1"),c("Estimate")], 3)`.
5.    Compared to corporations, two other business structures (partnership, with a coefficient of `r round(summary(step_interact_final)$coefficients[c("CorpStructurePartner"),c("Estimate")], 3)`, and limited liability corporation (LLC), with a coefficient of `r round(summary(step_interact_final)$coefficients[c("CorpStructureLLC"),c("Estimate")], 3)`)) are better. On the other hand, sole proprietorship has a negative but not statistically significant difference towards PRSM.
6.    The transformed Months variable (Months\^(`r lambda_months`)) gets smaller as the months a business is open increases. Thus its `r round(summary(step_interact_final)$coefficients[c("Months_transformed"),c("Estimate")], 3)` effect on PRSM decreases as the business is open longer, diminishing over time.

Interpretation of the interaction effects in the final model:

1.    **FICO:WomanOwned**. The negative effects of lower FICO categories increase if the business is women-owned, compared to businesses with an Excellent FICO score that are not woman-owned.
2.    **FICO:Months_transformed**. The positive effect of being open longer is less pronounced for businesses with FICO scores that belong to categories lower than Excellent. For the Poor FICO Score businesses, the effect of months is reduced by `r round(summary(step_interact_final)$coefficients[c("FICOPoor:Months_transformed"),c("Estimate")], 3)`. For the Fair FICO Score businesses, the effect of months is reduced by `r round(summary(step_interact_final)$coefficients[c("FICOFair:Months_transformed"),c("Estimate")], 3)`. For the Good FICO Score businesses, the effect of months is reduced by `r round(summary(step_interact_final)$coefficients[c("FICOGood:Months_transformed"),c("Estimate")], 3)`. And for the Very Good FICO Score businesses, the effect of months is reduced by `r round(summary(step_interact_final)$coefficients[c("FICOVery Good:Months_transformed"),c("Estimate")], 3)`.
3.    **Stress:WomanOwned**. The effect of Stress on PRSM is stronger by `r round(summary(step_interact_final)$coefficients[c("Stress:WomanOwned1"),c("Estimate")]*0.1, 3)` for women-owned business compared to other businesses.
4.    **WomanOwned:CorpStructure**.The LLC corporate structure does not perform as well if the business is owned by women by `r round(summary(step_interact_final)$coefficients[c("WomanOwned1:CorpStructureLLC"),c("Estimate")], 3)`, compared to corporations not owned by women. On the other hand, the partnership corporate structure performs better by `r round(summary(step_interact_final)$coefficients[c("WomanOwned1:CorpStructurePartner"),c("Estimate")], 3)`.
5.    **CorpStructure:Month_transformed**. For businesses structured as LLCs, the effect of months is more pronounced by an additional `r round(summary(step_interact_final)$coefficients[c("CorpStructureLLC:Months_transformed"),c("Estimate")], 3)`. For businesses structured as partnerships, the effect of months is even more steep by `r round(summary(step_interact_final)$coefficients[c("CorpStructurePartner:Months_transformed"),c("Estimate")], 3)`.

```{r baseline}
newdata<-installment2_id01_clean[1,]; print(newdata)
as.data.frame(round(predict(step_interact_final, 
        newdata=newdata), 8))
newdata[,c("FICO")] <- "Good"
newdata[,c("TotalAmtOwed")] <- median(installment2_id01_clean$TotalAmtOwed)
newdata[,c("Stress")] <- median(installment2_id01_clean$Stress)
newdata[,c("WomanOwned")] <- "0"
newdata[,c("CorpStructure")] <- "LLC"
newdata[,c("Months")] <- median(installment2_id01_clean$Months)
newdata[,c("Months_transformed")] <- median(installment2_id01_clean$Months)^lambda_months
print(newdata)
as.data.frame(round(predict(step_interact_final, 
        newdata=newdata), 8))
```

The baseline borrower has a Good FICO score, `r median(installment2_id01_clean$TotalAmtOwed)` of Total Amount Owed, and `r median(installment2_id01_clean$Stress)` of Stress, is not a woman, is structured as a LLC, and has opened the business for `r median(installment2_id01_clean$Months)` months.
