---
title: "STAT 628 Credit Risk Project: Installment 2 (technical)"
author: "Bohyoon Lee and Salsabila Mahdi"
date: last-modified
format: pdf
geometry:
  - margin=1in
---

```{r}
#| label: setup
#| echo: false
#| warning: false

rm(list = ls())

library(dplyr)
library(ModelMetrics)
library(MASS)
library(car)
library(caret)
library(glmnet)
library(ggplot2)
```

## Data Preprocessing

We load and preprocess the data to ensure the nonnumerical variables aren't 
considered numerical by R.

```{r load_clean}
installment2_id01 <- read.csv("installment2_id01.csv")
str(installment2_id01)
summary(installment2_id01)

# make NAICS, WomanOwned, FICO categorical variables

installment2_id01 = installment2_id01 %>%
  mutate(NAICS = as.factor(substr(NAICS, 1, 6)), # TODO: talk about simplifying
         WomanOwned = as.factor(WomanOwned),
         FICO = case_when(
           (300 <= FICO)&(FICO <= 579) ~ "Poor",
           (580 <= FICO)&(FICO <= 669) ~ "Fair",
           (670 <= FICO)&(FICO <= 739) ~ "Good",
           (740 <= FICO)&(FICO <= 799) ~ "Very Good",
           (800 <= FICO)&(FICO <= 850) ~ "Excellent",
         ))

```
We then notice a few points that are out of distribution and remove them.

```{r remove_outliers}
# TODO: rm by variable validity instead?
clean_PRSM <- subset(installment2_id01, PRSM > 0)

fullmodel <- lm(PRSM ~., data = installment2_id01)
plot(fullmodel,1)
installment2_id01$residual=residuals(fullmodel)

# remove 11 outliers observed in the plot (abs(residual)>=1)
installment2_id01_clean = installment2_id01 %>% 
  filter(abs(residual)<=1) %>% 
  dplyr::select(-residual) # clean data (use this!!)

model_clean = lm(PRSM ~. , data = installment2_id01_clean) 
plot(model_clean,1) # these look much better! but perhaps discuss different outlier removal techniques TODO: talk about this!

summary(model_clean)

```

## Model Development and Selection

### Linearity

We assess the linearity of the relationship between PRSM and the predictors 
using component-residual plots and find that Months shows nonlinearity in a somewhat logarithmic form. 
We transform Months to approximately its inverse (to be precise, Box-Tidwell suggests -1.1987).

```{r transform}
# Months seems to need transformation
crPlots(model_clean) 

# this matches hint 1 of the assignment
crPlots(lm(PRSM ~ Months, data = installment2_id01_clean))

# Transform Months Model
boxTidwell(PRSM ~ TotalAmtOwed + Volume + Stress + Num_Delinquent + Num_CreditLines + Months,
          data = installment2_id01_clean)

transformed_model <- lm(PRSM ~ . - Months + I(Months^(-1.19487)),
                       data = installment2_id01_clean)
# TODO: don't hardcode

summary(transformed_model)
rmse(transformed_model$fitted.values, installment2_id01_clean$PRSM) # 0.1018195

crPlots(lm(PRSM ~ I(Months^(-1.19487)), data = installment2_id01_clean))
```
### Variable and Interaction Selection

We aim to create a parsimonious model that explains the most variance. 
We do this in a multi-stage selection process.

First, we use bidirectional stepwise on the model containing all main effects (with Months transformed). 
It identifies NAICS, Num_Delinquent, and Num_CreditLines as non-significant predictors and removes them.
We also remove Volume as it is non-significant and its effect is most likely covered by Stress.

```{r rm_var}
# TODO: talk about just doing step instead of LASSO and Step?
# Step Model
step_model <- step(transformed_model, trace = 0, direction = "both")
summary(step_model)
# Step Model removed NAICS, Num_Delinquent, and Num_CreditLines
rmse(step_model$fitted.values, installment2_id01_clean$PRSM) # RMSE = 0.1025071


# 10-fold Cross-Validation
cv10 = trainControl(method = "cv", number = 10)
model_cv = train(PRSM ~ FICO + TotalAmtOwed + Volume + Stress + WomanOwned + 
      CorpStructure + I(Months^(-1.19487)), data = installment2_id01_clean, 
      method ="lm", trControl =cv10) # RMSE = 0.1031503
summary(model_cv)

# Volume is not significant (remove later)
```
Secondly, we search for statistically significant two-way interactions with LASSO, that affect PRSM meaningfully. 
We add quite a few interactions between FICO, Stress, WomanOwned and CorpStructure with other predictors to the model.

```{r interaction_lasso}

# Find interactions using LASSO
installment2_id01_clean$Months_transformed =
  (installment2_id01_clean$Months)^(-1.19487)
x <- model.matrix(PRSM ~ .^2, 
                  data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed")])[, -1]
y <- installment2_id01_clean$PRSM
cv_lasso_interactions <- cv.glmnet(x, y, alpha = 1, nfolds = 10)

best_lambda <- cv_lasso_interactions$lambda.min
lasso_coeffs <- coef(cv_lasso_interactions, s = best_lambda)

threshold <- 0.001
significant_terms <- lasso_coeffs[abs(lasso_coeffs[,1]) > threshold, , drop = FALSE]
print(significant_terms)

# potential interactions: FICO:Stress, FICO:WomanOwned, FICO:CorpStructure, FICO:Months_transformed, Stress:WomanOwned, Stress:CorpStructure, WomanOwned:CorpStructure, WomanOwned:Months_transformed, CorpStructure:Months_transformed

interact_model = lm(PRSM ~. + FICO:Stress + FICO:WomanOwned + FICO:CorpStructure 
                    + FICO:Months_transformed + Stress:WomanOwned 
                    + Stress:CorpStructure + WomanOwned:CorpStructure 
                    + WomanOwned:Months_transformed + CorpStructure:Months_transformed
                      , data = installment2_id01_clean[,c("PRSM", "FICO",
                                                    "TotalAmtOwed","Stress",
                                                    "WomanOwned", "CorpStructure",
                                                    "Months_transformed")])
summary(interact_model)
step_interact_model = step(interact_model, trace = 0, direction = "both")
summary(step_interact_model) # printed R summary of the fitted model
# Step Model removed NAICS, Num_Delinquent, and Num_CreditLines
rmse(step_interact_model$fitted.values, installment2_id01_clean$PRSM) # RMSE = 0.09830548


# TODO: look over hints..., multiple regression assumptions, interpretations of the parameters, explanation for how we selected the characteristics of the baseline borrower

# CHECK ASSUMPTIONS FOR FINAL MODEL
plot(step_interact_model, 2)
plot(step_interact_model, 1)
vif(step_interact_model) # <5, still ok for multicol

# current findings:
# Stress:WomanOwned1, and LLC but smaller => pos. 
# WomanOwned1:CorpStructurePartner => pos.
# FICOFair&Poor:WomanOwned1 => neg!
# FICOFair&Good:Monts => neg. Hint diminished effect again
# CorpStructureSole:p_delinquent => neg. p-delinquent harder on Sole owner


ggplot(data = installment2_id01_clean, aes(x = WomanOwned)) +
  geom_bar() +
  facet_wrap(~ FICO)

```

Finally, we test the validity of the model, with variables properly casted and
outliers removed from the data, Months transformed, selected main effects, and interaction terms.
We find the model to meet the assumptions and have the lowest RMSE so we choose it as the final model.
